### 2015-05-12 whoop.tex
fixes

\medskip\noindent\textbf{Lockset instrumentation }
%
Suppose that $S$ and $T$ are thread templates for a concurrent program, including the possibly that $S=T$.
We wish to check whether it is possible for a thread that is an instance of $S$ to race with a thread that is
an instance of $T$, in the presence of arbitrarily many further threads.

The first phase in achieving this is to \emph{instrument} a thread template to enable lockset analysis.  Given an arbitrary symbol, $i$, we define the instrumentation of a
template $T$ with respect to $i$, denoted $\instrument{T}{i}$.  There are two aspects to instrumentation, \emph{renaming} and \emph{lockset instrumentation}.

Renaming is simple: every private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.  The purpose of renaming is to ensure that when we analyse a pair of templates, $S$ and $T$, we ensure that $S$ and $T$ execute distinct procedures and operate on distinct private data.  Shared variables and mutexes are \emph{not} renamed, because these are indeed shared across templates.

@@

$\mutexunlock{m};$ & $\currentlockset{i} = \currentlockset{i} \setminus \{ m \};$ \\
\bottomrule
\end{tabular}
\caption{Instrumenting statements for static lockset analysis}\label{fig:instrumentation}
\end{figure}

Lockset instrumentation first requires the introduction of sets to track those variables that have been read from and written to, the mutexes that have been consistently used to protect each variable, and the set of current locks held by a thread.  To this end, when instrumenting template $T$ with respect to symbol $i$, we add a sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track those shared variables that have been read from and written to, respectively, by the thread executing $T$.  We add a current lock set, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread, and for each shared variable $s \in \sharedvars$ we add a lock set, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.

@@

\medskip\noindent\textbf{Shared state abstraction }
Recall that while our aim is to perform race analysis for pairs of threads, we must be sure to account
for possible side-effects due to other threads that are running concurrently.  The instrumentation
of Figure~\ref{fig:instrumentation} achieves this via \emph{nondeterminism}: when reading from a shared variable $s$,
a nondeterministic value is returned.  This is reflected by the use of a $\havockeyword$ command,
which sets its argument to an arbitrary value.  Because all shared state accesses are abstracted in
this fashion, it is possible to completely dispense with the shared variables after instrumentation
has been performed.  As a result, when instrumenting a shared variable write, the effect of the write
is not explicitly modelled.

\medskip\noindent\textbf{Sequentialisation }
The psuedocode of Figure~\ref{fig:sequentialization} shows the sequential program that we
analyse in order to prove race-freedom for a pair of thread templates $T$ and $U$.
Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding
$\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows.
First, the read, write and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are
initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and
$\lockset{s}{j}$ are initialized to the full set of mutexes, $\mutexes$.  The main procedures
of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and
$\mainforthread{\instrument{U}{j}}$ are then executed in turn.  Finally, an assertion checks
for consistent use of mutexes: if a shared variables is written during execution of $\instrument{T}{i}$
and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$
and $\lockset{s}{j}$ must contain a common mutex.

@@

\medskip\noindent\textbf{Soundness }
We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct,
it is impossible for a thread executing template $T$ to race with a thread executing template $U$,
under the assumption that the threads are guaranteed to terminate.
Let us assume that the program of Figure~\ref{fig:sequentialization} is correct,
and suppose (by way of contradition) that a thread executing $T$ can in fact race with a thread
executing $U$, on some shared variable $s$.  By our hypothesis that the program
of Figure~\ref{fig:sequentialization} is correct, and that the threads terminate,
the assertion checked at the end of the program guarantees at least one mutex, $m$ say,
belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset
(and according to the manner in which shared accesses are instrumented in Figure~\ref{ }),
this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$.
As a result, $m$ must be unlocked and locked between the two accesses, which
contradicts that the pair of accesses is racing.

>>>>

\medskip\noindent\textbf{Lockset instrumentation }
%
Suppose that $S$ and $T$ are thread templates for a concurrent program, including the possibility that $S$ and $T$ are equal. We wish to check whether it is possible for a thread that is an instance of $S$ to race with a thread that is an instance of $T$, in the presence of arbitrarily many further threads that may execute concurrently.

To achieve this, we first \emph{instrument} a thread template for lockset analysis.  Given an arbitrary symbol, $i$, we define the instrumentation of a template $T$ with respect to $i$, denoted $\instrument{T}{i}$.  There are two aspects to this instrumentation phase: \emph{renaming} and \emph{lockset instrumentation}.

Renaming is straightforward: every private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.  The purpose of renaming is to ensure that when we analyze a pair of templates, $S$ and $T$, both templates execute distinct procedures and operate on distinct private data.  Shared variables and mutexes are \emph{not} renamed, because these are shared across templates.

@@

$\mutexunlock{m};$ & $\currentlockset{i} = \currentlockset{i} \setminus \{ m \};$ \\
\bottomrule
\end{tabular}
\caption{Instrumenting statements for lockset analysis}\label{fig:instrumentation}
\end{figure}

Lockset instrumentation first requires the introduction of sets to track those variables that have been read from and written to, the mutexes that have been consistently used to protect each variable, and the set of current locks held by a thread.  To this end, when instrumenting template $T$ with respect to symbol $i$, we add sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track those shared variables that have been read from and written to, respectively, by the thread executing $T$.  We add a current lock set, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread, and for each shared variable $s \in \sharedvars$ we add a lock set, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.

@@

\medskip\noindent\textbf{Shared state abstraction }
%
Recall that while our aim is to perform race analysis for pairs of threads, we must be sure to account
for possible side-effects due to other threads that are running concurrently.  The instrumentation
of Figure~\ref{fig:instrumentation} achieves this via \emph{nondeterminism}: when reading from a shared variable $s$,
a nondeterministic value is returned.  This is reflected by the use of a $\havockeyword$ command,
which sets its argument to an arbitrary value.  Because all shared state accesses are abstracted in
this fashion, it is possible to completely dispense with the shared variables after instrumentation
has been performed.  As a result, when instrumenting a shared variable write, the effect of the write
is not explicitly modeled.

\medskip\noindent\textbf{Sequentialisation }
%
The pseudocode of Figure~\ref{fig:sequentialization} shows the sequential program that we
analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$.
Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding
$\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows.
First, the read, write and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are
initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and
$\lockset{s}{j}$ are initialized to the full set of mutexes, $\mutexes$.  The main procedures
of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and
$\mainforthread{\instrument{U}{j}}$ are then executed in turn.  Finally, an assertion checks
for consistent use of mutexes: if a shared variables is written during execution of $\instrument{T}{i}$
and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$
and $\lockset{s}{j}$ must contain a common mutex.

@@

\medskip\noindent\textbf{Soundness }
%
We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct,
it is impossible for a thread executing template $T$ to race with a thread executing template $U$,
under the assumption that the threads are guaranteed to terminate.
Let us assume that the program of Figure~\ref{fig:sequentialization} is correct,
and suppose (by way of contradition) that a thread executing $T$ can in fact race with a thread
executing $U$, on some shared variable $s$.  By our hypothesis that the program
of Figure~\ref{fig:sequentialization} is correct, and that the threads terminate,
the assertion checked at the end of the program guarantees at least one mutex, $m$ say,
belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset
(and according to the manner in which shared accesses are instrumented in Figure~\ref{ }),
this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$.
As a result, $m$ must be unlocked and locked between the two accesses, which
contradicts that the pair of accesses is racing.
